Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-11-22 source-iclr/sec_Nesterov.tex
Update on Overleaf.

A natural extension of the FedAvg algorithm is to use momentum-based
local updates instead of local SGD updates. To our knowledge, the
only convergence analyses of FedAvg with momentum-based stochastic
updates focus on the non-convex smooth case \cite{huo2020faster,yu2019linear,li2018federated}.
In this section, we complete the picture by providing the first $\mathcal{O}(1/KT)$
and $\mathcal{O}(1/\sqrt{KT})$ convergence results for Nesterov-accelerated
FedAvg for convex objectives that match the rates from the previous section. Detailed proofs of convergence results in this section are deferred to Appendix Section~\ref{sec:app:Nesterovfedavg}.

@@

We now show that the optimality gap of Nesterov Accelerated FedAvg has $\mathcal{O}(1/\sqrt{KT})$ rate for convex and smooth objectives. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{KT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:boundedvariance},\ref{ass:subgrad2} Nesterov accelerated FedAvg with
	full device participation has rate
	% \vspace{-0em}
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right),
		\end{align*}}
		% \vspace{-1em}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right).
		\end{align*}}
\end{theorem}
%
	It is possible to extend the results in this section to accelerated
    FedAvg algorithms with other momentum-based updates. However, as know from stochastic optimization, Nesterov and other momentum
updates may fail to accelerate over SGD (\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}).
    For this reason, we will instead turn to the overparameterized setting
    \cite{ma2017power,liu2018accelerating,canziani2016analysis} in Section~\ref{sec:app:overparameterized} of the appendix where we show that FedAvg enjoys geometric convergence
    and it is possible to improve its convergence rate with a new momentum-based FedAvg variant, which we term ``FedMaSS''.

>>>>

{\color{red}A natural extension of the FedAvg algorithm is to use momentum-based 
local updates instead of local SGD updates in order to accelerate FedAvg. As we know from stochastic optimization, Nesterov and other momentum
updates fail to provably accelerate over SGD (\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2020federated}). This is in contrast to the classical acceleration result of Nesterov-accelerated gradient descent over GD. Thus in the FL setting, the best provable convergence rate for FedAvg with Nesterov updates is the same as FedAvg with SGD updates. Nevertheless, Nesterov and other momentum updates are frequently used in practice, in both non-FL and FL settings, and are observed to perform better empirically. In fact, previous works such as \cite{stich2018local} on FedAvg with vanilla SGD uses FedAvg with Nesterov or other momentum updates in their experiments to achieve target accuracy. Because of the popularity of Nesterov and other momentum-based methods, understanding the linear speedup behavior of FedAvg with such local updates is important.} To our knowledge, the
only convergence analyses of FedAvg with momentum-based stochastic
updates focus on the non-convex smooth case \cite{huo2020faster,yu2019linear,li2018federated}, and no results existed in the convex smooth settings. In this section, we complete the picture by providing the first $\mathcal{O}(1/KT)$
and $\mathcal{O}(1/\sqrt{KT})$ convergence results for Nesterov-accelerated
FedAvg for convex objectives that match the rates for FedAvg with SGD updates. Detailed proofs of convergence results in this section are deferred to Appendix Section~\ref{sec:app:Nesterovfedavg}.

@@

We now show that the optimality gap of Nesterov-accelerated FedAvg has $\mathcal{O}(1/\sqrt{KT})$ rate for convex and smooth objectives. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{KT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:boundedvariance},\ref{ass:subgrad2} Nesterov accelerated FedAvg with
	full device participation has rate
	% \vspace{-0em}
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right),
		\end{align*}}
		% \vspace{-1em}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\small{\begin{align*}
		\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right).
		\end{align*}}
\end{theorem}

We emphasize again that in the stochastic optimization setting, the optimal convergence rate that FedAvg wtih Nesterov udpates can achieve is the same as FedAvg with SGD updates. However, due to the popularity and superior performance of momentum methods in practice, it is still important to understand the linear speedup behavior of such FedAvg variants. Our results in this section fill exactly this gap.

	\begin{comment}
	It is possible to extend the results in this section to accelerated
    FedAvg algorithms with other momentum-based updates. However, as know from stochastic optimization, Nesterov and other momentum
updates may fail to accelerate over SGD (\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}).
    For this reason, we instead turn to the overparameterized setting
    \cite{ma2017power,liu2018accelerating,canziani2016analysis} in Section~\ref{sec:app:overparameterized} of the appendix where we show that FedAvg enjoys geometric convergence
    and it is possible to improve its convergence rate with a new momentum-based FedAvg variant, which we term ``FedMaSS''.
	\end{comment}

### 2020-11-22 source-iclr/sec_intro.tex
Update on Overleaf.

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated (\cite{kairouz2019advances}), it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization (\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified}):

@@

\textbf{Connections with Distributed and Decentralized Optimization}
Federated learning is closely related to distributed and decentralized optimization, and as such it is important to discuss connections and distinctions between our work and related results from that literature. First, when there is neither system heterogeneity, i.e. all devices participate in parameter averaging during a communication round, nor statistical heterogeneity, i.e. all devices have access to a common set of stochastic gradients, FedAvg coincides with the ``Local SGD'' of~\cite{stich2018local}, which showed the linear speedup rate $\mathcal{O}(1/NT)$ for strongly convex and smooth functions. When there is only data heterogeneity, some works have continued to use the term Local SGD to refer to FedAvg, while others subsume it in more general frameworks that include decentralized model averaging based on a network topology or a mixing matrix. They have provided linear speedup analyses for strongly convex and convex problems, e.g.~\cite{khaled2020tighter,koloskova2020unified} as well as non-convex problems, e.g.~\cite{jiang2018linear,yu2019parallel,wang2018cooperative}. However, these results do not consider system heterogeneity, i.e. the presence of stragglers in the device network. Even with decentralized model averaging, the assumptions usually imply that model averages over all devices is the same as decentralized model averages based on network topology (e.g.~\cite{koloskova2020unified} Proposition 1), which precludes system heterogeneity as defined in this paper and prevalent in FL problems. For momentum accelerated FedAvg,~\cite{yu2019linear} provided linear speedup analysis for non-convex problems, while results for strongly convex and convex settings are entirely lacking, even without system heterogeneity.~\cite{karimireddy2019scaffold} considers both types of heterogeneities, but their rate implies a linear speedup only when the number of stragglers is negligible. In contrast, our linear speedup analyses consider both types of heterogeneity present in the full federated learning setting, and are valid for any number of participating devices. We also highlight a distinction in communication efficiency when system heterogeneity is present. Moreover, our results for Nesterov accelerated FedAvg completes the picture for strongly convex and convex problems. For a detailed comparison with related works, please refer to Table~\ref{tb:convergenceratev3} in Appendix Section~\ref{sec:app:comparison}.

>>>>

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated (\cite{kairouz2019advances}), it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization (\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified,woodworth2020local,woodworth2020minibatch}):

@@

\textbf{Connections with Distributed and Decentralized Optimization}
Federated learning is closely related to distributed and decentralized optimization, and as such it is important to discuss connections and distinctions between our work and related results from that literature. First, when there is neither system heterogeneity, i.e. all devices participate in parameter averaging during a communication round, nor statistical heterogeneity, i.e. all devices have access to a common set of stochastic gradients, FedAvg coincides with the ``Local SGD'' of~\cite{stich2018local}, which showed the linear speedup rate $\mathcal{O}(1/NT)$ for strongly convex and smooth functions. \cite{woodworth2020local} and \cite{woodworth2020minibatch} further improved the communication complexity that guarantees the linear speedup rate. When there is only data heterogeneity, some works have continued to use the term Local SGD to refer to FedAvg, while others subsume it in more general frameworks that include decentralized model averaging based on a network topology or a mixing matrix. They have provided linear speedup analyses for strongly convex and convex problems, e.g.~\cite{khaled2020tighter,koloskova2020unified} as well as non-convex problems, e.g.~\cite{jiang2018linear,yu2019parallel,wang2018cooperative}. However, these results do not consider system heterogeneity, i.e. the presence of stragglers in the device network. Even with decentralized model averaging, the assumptions usually imply that model averages over all devices is the same as decentralized model averages based on network topology (e.g.~\cite{koloskova2020unified} Proposition 1), which precludes system heterogeneity as defined in this paper and prevalent in FL problems. For momentum accelerated FedAvg,~\cite{yu2019linear} provided linear speedup analysis for non-convex problems, while results for strongly convex and convex settings are entirely lacking, even without system heterogeneity.~\cite{karimireddy2019scaffold} considers both types of heterogeneities for FedAvg, but their rate implies a linear speedup only when the number of stragglers is negligible. In contrast, our linear speedup analyses consider both types of heterogeneity present in the full federated learning setting, and are valid for any number of participating devices. We also highlight a distinction in communication efficiency when system heterogeneity is present. Moreover, our results for Nesterov accelerated FedAvg completes the picture for strongly convex and convex problems. For a detailed comparison with related works, please refer to Table~\ref{tb:convergenceratev3} in Appendix Section~\ref{sec:app:comparison}.
